Report on Late Winter2013 Production: Image Differencing

### 2013-03-22 W13report.tex
Fixed table with false positives based on Simons association.  Added backgrund fitting description.

We assume that science image $S(x,y)$ can be modeled as a convolution
of the template image $T(x,y)$ by a PSF--matching kernel $K(u,v,x,y)$.
We further assume that the kernel may be decomposed using a set of
basis functions $K(u,v) = \sum_i a_i K_i(u,v)$, where the coefficients
in front of each basis are determined through:
%
\begin{eqnarray}
C_i & \equiv & (K_i \otimes T) \\ \nonumber
b_{i}  & = & \sum_{x,y} {{C_i(x,y) S(x,y)}\over{\sigma^2(x,y)}}   \nonumber \\ 
M_{ij} & = & \sum_{x,y} {{C_i(x,y) C_j(x,y)}\over{\sigma^2(x,y)}}  \nonumber \\ 
a_{i}  & = & M^{-1}_{ij} b_{j} \nonumber 
\label{eq-soln}
\end{eqnarray}
where $\sigma^2(x,y)$ is the per--pixel variance stored in the {\tt
  variance} plane of each LSST {\tt exposure}.  To generate a
spatially varying model for the kernel, we assume that the relative
weights of the basis functions $a_i$ themselves vary spatially,
i.e. $K(u,v,x,y) = \sum_i a_i(x,y) K_i(u,v)$.  The image difference is
then calculated through $D(x,y) = S(x,y) - T(x,y) \otimes K(u,v,x,y)$.

@@

The coadd and science {\tt calexps} are then astrometrically
registered to each other, with the resampling operations happening on
the coadd as it is expected to have larger S/N.  Object--to--object
matching, done using the source list above cross--matched with a
source list extracted from the coadd, is used to provide a relative
transformation between the two images.  RegisterTask is used to do the
astrometric resampling.  The resulting warped coadd is what we use as
the template image $T(x,y)$.  The residuals of the object--to--object
match are persisted as part of the debugging infrastructure.

@@

All {\tt DiaSources} are associated with the {\tt calexp's} {\tt
  sources} and with the reference catalog, with a permissive matching
radius of 2''.  {\tt DiaSources} that match with {\tt sources} may be
true residuals around stars {\it or} true false detections that are
also found in {\tt calexp} SourceDetectionTask.  For this reason, we
use the associations with the reference catalog objects as the
assessment if a given {\tt DiaSource} matches with a {\tt source} or
whether it is an orphan (e.g. a statistical fluctuation in the
background).  In addition, all {\tt KernelCandidates} are persisted
with the above {\tt LOCAL} and {\tt SPATIAL} metrics.  Repositories
are accessed with a specially written RepositoryIterator that
aggregates {\tt KernelCandidates} from a given {\tt source} across
multiple output repositories.

@@

@@

After running an ensemble of 144 pipeline configurations, we
aggregated the numbers of false detections.  We report the results for
the each of the 3 visits, and for pre-- and postfiltering for
detection.  We cut on {\tt flags.pixel.edge} to remove spurious
detections on the immediate border of the image; other than that we do
no other cuts other than requiring that the {\tt centroid.sdss} is
finite.  We count the DiaSources that match with the reference catalog
(2'' matching radius) via the {\tt refMatchId} value.  We find no
significant correlation of false detections with sensor within the
raft; therefore we report the aggregate results for all 9 sensors in
the analysis that follows.

@@

The numbers of false detections in each best case are given in
Table~\ref{tab-bestfp10}.  We undertook a manual inspection of all
difference images to associate the DiaSources with Sources in the
images.  These by--eye associations are listed in the Matches column
of Table~\ref{tab-bestfp10}, while the remainder are listed in the
Orphans column.  A significant number of the Matches (e.g. both the
matches for v6866603) are actually DiaSources on the very edge of the
image, but which do not have the EDGE bit set.

indicating that the residuals are all likely fluctuations in the
backgrounds.

@@

The variance in the prefiltered and postfiltered images is likely to be
different due to the different types of operations that have happened
on the images.  We quantify this by taking the postfiltered images,
immediately before the detection step (i.e. $D'$), and compare the
variance plane to that coming from the prefiltered image.  For
v6866601, which is a deconvolution in the postfiltered production, we
find that the variance is 70\% higher when postfiltering.  This is
consistent with the expectation that deconvolution increases the noise
properties of the image.

For v6866602 and v6866603, we find that the variance is 3--4\% and
2--3\% higher in the {\it prefiltered} data, respectively.  The
primary culprit for this is that the prefiltered data will have gone
through 1 less convolution, meaning less of the variance will have
been lost to covariance.  In this case, the variance plane will be
closer to the truth than in the postfiltered data, but still not
exact.  Assuming that most of the covariance happens during
Psf--matching and Psf--filtering (the astrometric warping kernel is
designed to preserve the noise properties of the image, although this
is not an exact conservation in practice), the variance plane of the
prefiltered data should itself be $\sim$3\% lower than the empirical
variance in the image plane.  {\it TEST ME}.

@@

\clearpage
\begin{table}
\centering
\begin{tabular}{clcc}
\hline
\multicolumn{4}{|c|}{Best Results: Production 8} \\
\hline
Visit    & Prefilter & False Detection Matches & False Detection Orphans \\
\hline
v6866601 & True      & 5                      & 29  \\
         & False     & 10                     & 81  \\
v6866602 & True      & 5                      & 30  \\
         & False     & 9                      & 35  \\
v6866603 & True      & 2                      & 9   \\
         & False     & 1                      & 23  \\
\end{tabular}
\caption{Total number of {\tt DiaSources} detected at 5--sigma from
  all 9 sensors in raft 2,2, for the configurations that yielded the
  lowest number of false detections.  Matches are determined using a
  2'' match radius with the input reference catalog.  Objects with the
  {\tt EDGE} flag are ignored. \label{tab-bestfp8}}
\end{table}

\begin{table}
\centering
\begin{tabular}{clccc}
\hline
\multicolumn{5}{|c|}{Best Results: Production 10} \\
\hline
Visit    & Prefilter & False Detections: Matches & False Detections: Orphans & Nneg / Npos \\
\hline
v6866601 & True      & 11                        & 59  & 1.8 \\ 
         & False     & 99                        & 44  & 2.9 \\
v6866602 & True      & 7                         & 35  & 2.2 \\
         & False     & 8                         & 45  & 0.7 \\
v6866603 & True      & 1                         & 22  & 2.8 \\
         & False     & 1                         & 35  & 3.0 \\
%v6866601 & True      & 11                        & 70  & 1.8 \\ SUBTRACT ME!
%         & False     & 99                        & 143 & 2.9 \\
%v6866602 & True      & 7                         & 42  & 2.2 \\
%         & False     & 8                         & 53  & 0.7 \\
%v6866603 & True      & 1                         & 23  & 2.8 \\
%         & False     & 1                         & 36  & 3.0 \\
\end{tabular}
\caption{Total number of {\tt DiaSources} detected at 5--sigma from
  all 9 sensors in raft 2,2, for the configurations that yielded the
  lowest number of false detections.  Matches are determined using a
  2'' match radius with the input reference catalog.  Objects with the
  {\tt EDGE} flag are ignored.  The final column lists ratios of the
  number of negatively detected false detections to those with
  positive flux. \label{tab-bestfp10}}
\end{table}

\begin{table}
\centering
\begin{tabular}{clccc}
\hline
\multicolumn{5}{|c|}{Best Configs: Production 8} \\
\hline
Visit    & Prefilter & nGauss & degGauss & spatialOrder \\
\hline
v6866601 & True      & 3      & 6        & 5 \\
         & False     & 3      & 2        & 4 \\
v6866602 & True      & 3      & 6        & 5 \\
         & False     & 3      & 5        & 5 \\
v6866603 & True      & 3      & 6        & 4 \\
         & False     & 3      & 6        & 4 \\
\end{tabular}
\caption{Configurations that led to the best results listed in
  Table~\ref{tab-bestfp8}.  All post-production runs include
  perturbations about these configurations. \label{tab-bestconfig8}}
\end{table}

>>>>

We assume that science image $S(x,y)$ can be modeled as a convolution
of the template image $T(x,y)$ by a PSF--matching kernel $K(u,v,x,y)$.
We further assume that the kernel may be decomposed using a set of
basis functions $K(u,v) = \sum_i a_i K_i(u,v)$, where the coefficients
in front of each basis are determined through:
%
\begin{eqnarray}
C_i & \equiv & (K_i \otimes T) \\ \nonumber
b_{i}  & = & \sum_{x,y} {{C_i(x,y) S(x,y)}\over{\sigma^2(x,y)}}   \nonumber \\ 
M_{ij} & = & \sum_{x,y} {{C_i(x,y) C_j(x,y)}\over{\sigma^2(x,y)}}  \nonumber \\ 
a_{i}  & = & M^{-1}_{ij} b_{j} \nonumber 
\label{eq-soln}
\end{eqnarray}
where $\sigma^2(x,y)$ is the per--pixel variance stored in the {\tt
  variance} plane of each LSST {\tt exposure}.  To generate a
spatially varying model for the kernel, we assume that the relative
weights of the basis functions $a_i$ themselves vary spatially,
i.e. $K(u,v,x,y) = \sum_i a_i(x,y) K_i(u,v)$.  We also assume that
there is a differnetial background between the two images $b(x,y)$
that may be fit for using a low--order Chebyshev.  The image
difference is then calculated through $D(x,y) = S(x,y) - T(x,y)
\otimes K(u,v,x,y) - b(x,y)$.

@@

The coadd and science {\tt calexps} are then astrometrically
registered to each other, with the resampling operations happening on
the coadd as it is expected to have larger S/N.  Object--to--object
matching, done using the source list above cross--matched with a
source list extracted from the coadd, is used to provide a relative
transformation between the two images.  RegisterTask is used to do the
astrometric resampling.  The resulting warped coadd is what we use as
the template image $T(x,y)$.  The residuals of the object--to--object
match are persisted as part of the debugging infrastructure.  We
assume that the background subtraction on the input {\tt calexps} has
removed the high order features in the backgrounds; therefore we
simultaneously fit for a 1st--order spatially varying differential
background $b(x,y)$ along with the kernel coefficients.

@@

All {\tt DiaSources} are associated with the {\tt calexp's} {\tt
  sources} and with the reference catalog, with a permissive matching
radius of 3''.  {\tt DiaSources} that match with {\tt sources} may be
true residuals around stars {\it or} true false detections that are
also found in {\tt calexp} SourceDetectionTask.  For this reason, we
use the associations with the reference catalog objects as the
assessment if a given {\tt DiaSource} matches with a {\tt source} or
whether it is an orphan (e.g. a statistical fluctuation in the
background).  In addition, all {\tt KernelCandidates} are persisted
with the above {\tt LOCAL} and {\tt SPATIAL} metrics.  Repositories
are accessed with a specially written RepositoryIterator that
aggregates {\tt KernelCandidates} from a given {\tt source} across
multiple output repositories.

@@

\subsection{Control Sample {\bf SIMON}}

@@

After running an ensemble of 144 pipeline configurations, we
aggregated the numbers of false detections.  We report the results for
the each of the 3 visits, and for pre-- and postfiltering for
detection.  We cut on {\tt flags.pixel.edge} to remove spurious
detections on the immediate border of the image; other than that we do
no other cuts other than requiring that the {\tt centroid.sdss} is
finite.  We count the DiaSources that match with the reference catalog
(3'' matching radius) via the {\tt refMatchId} value.  We find no
significant correlation of false detections with sensor within the
raft; therefore we report the aggregate results for all 9 sensors in
the analysis that follows.

@@

The numbers of false detections in each best case are given in
Table~\ref{tab-bestfp10a}.  We undertook a manual inspection of all
difference images to associate the DiaSources with Sources in the
images, and also a 3'' association with the input source catalog.
These associations are listed in the N$_{Match}$ column of
Table~\ref{tab-bestfp10}, with the unmatched false positives in the
N$_{Orphan}$ column.  The former are likely to come from systematics
in the subtractions of stars, while the latter would be due to
fluctuations in the background.  We note 2 effects that will impact
N$_{Match}$.  The first is that, with a 3'' matching radius, 4\% of
random fluctuations will by--chance associate with a Source.  Assuming
that N$_{Orphan}$ represents 96\% of the background fluctuation
population, we estimate the number of fluctuations from that same
population that would have associated with Sources as N$_{Random}$.
Finally, we count all matches that occur within 100 pixels of the
image boundary, where we might expect the spatial extrapolation to
yield systematic residuals, as N$_{D<100}$.  To the extent that
(N$_{Random}$ + N$_{D<100}$) < N$_{Match}$, we have a small residual
set of systematic false detections in our imaging ensemble.  Aside
from the deconvolution run, this is at the level of 10 total false
detections over 9 sensors in 5 production runs, or 1 every 4--5
images.

@@

The variance in the prefiltered and postfiltered images is likely to
be different due to the different types of operations that have
happened on the images.  We quantify this by taking the postfiltered
images, immediately before the detection step (i.e. $D'$), and compare
the variance plane to that coming from the prefiltered image
(Table~\ref{tab-variance2}).  For v6866601, which is a deconvolution
in the postfiltered production, we find that the variance is 70\%
higher when postfiltering.  This is consistent with the expectation
that deconvolution increases the noise properties of the image.  For
v6866602 and v6866603, we find that the variance is 3--4\% and 2--3\%
higher in the {\it prefiltered} data, respectively.  The primary
culprit for this is that the prefiltered data will have gone through 1
less convolution, meaning less of the variance will have been lost to
covariance.

To quantify the significance of these offsets, we compared the
empirical variance in the image planes with the variance tracked by
the variance planes to ascertain how much we are misestimating the
true variance.  We calculated the ratio of the interquartile range of
unmasked pixels in the image plane, multiplied by 0.741 and then
squared, with the median of unmasked pixels from the variance plane.
The results are summarized in Table~\ref{tab-variance1}.  We find that
we are marginally, but consistently, underestimating the variance in
our images by $\sim 1\%$.  This affect is smaller than the amplitudes
reported above, indicating that there are true differences in the
variances of prefiltered and postfiltered detection images.

@@

\clearpage
%\begin{table}
%\centering
%\begin{tabular}{clcc}
%\hline
%\multicolumn{4}{|c|}{Best Results: Production 8} \\
%\hline
%Visit    & Prefilter & False Detection Matches & False Detection Orphans \\
%\hline
%v6866601 & True      & 5                      & 29  \\
%         & False     & 10                     & 81  \\
%v6866602 & True      & 5                      & 30  \\
%         & False     & 9                      & 35  \\
%v6866603 & True      & 2                      & 9   \\
%         & False     & 1                      & 23  \\
%\end{tabular}
%\caption{Total number of {\tt DiaSources} detected at 5--sigma from
%  all 9 sensors in raft 2,2, for the configurations that yielded the
%  lowest number of false detections.  Matches are determined using a
%  3'' match radius with the input reference catalog.  Objects with the
%  {\tt EDGE} flag are ignored. \label{tab-bestfp8}}
%\end{table}

\begin{table}
\centering
\begin{tabular}{clcccccc}
\hline
\multicolumn{8}{|c|}{Best Results: Production 10} \\
\hline
Visit    & Prefilter & Total False Detections &  N$_{Orphan}$ & N$_{Match}$ & N$_{Random}$ & N$_{D<100}$ & Nneg / Npos \\
\hline
v6866601 & True      & 70      &60         & 10 & 3     & 4   & 1.8 \\ 
         & False     & 143     &50         & 93 & 2     & 11  & 2.9 \\
v6866602 & True      & 42      &36         & 6  & 2     & 1   & 2.2 \\
         & False     & 53      &45         & 8  & 2     & 1   & 0.7 \\
v6866603 & True      & 23      &22         & 1  & 1     & 1   & 2.8 \\
         & False     & 36      &35         & 1  & 1     & 0   & 3.0 \\
\end{tabular}
\caption{Total number of {\tt DiaSources} detected at 5--sigma from
  all 9 sensors in raft 2,2, for the configurations that yielded the
  lowest number of false detections.  Matches are determined using a
  3'' match radius with the input reference catalog.  Objects with the
  {\tt EDGE} flag are ignored.  We list the number of background
  fluctuations that are expected to randomly associate with a Source
  N$_{Random}$, given the density of objects in the images and a 3''
  match radius, assuming that N$_{Orphan}$ represents 96\% of the
  total population of detections arriving from Gaussian fluctuations.
  We also list the number of false detections that are found within
  the outer 100 pixels of the image N$_{D<100}$.  The final column
  lists ratios of the number of negatively detected false detections
  to those with positive flux, a 5--sigma. \label{tab-bestfp10a}}
\end{table}

\begin{table}
\centering
\begin{tabular}{clccc}
\hline
\multicolumn{5}{|c|}{Best Results: Production 10} \\
\hline
Visit    & Prefilter & False Detections: Matches & False Detections: Orphans & Nneg / Npos \\
\hline
v6866601 & True      & 11                        & 59  & 1.8 \\ 
         & False     & 99                        & 44  & 2.9 \\
v6866602 & True      & 7                         & 35  & 2.2 \\
         & False     & 8                         & 45  & 0.7 \\
v6866603 & True      & 1                         & 22  & 2.8 \\
         & False     & 1                         & 35  & 3.0 \\
%v6866601 & True      & 11                        & 70  & 1.8 \\ SUBTRACT ME!
%         & False     & 99                        & 143 & 2.9 \\
%v6866602 & True      & 7                         & 42  & 2.2 \\
%         & False     & 8                         & 53  & 0.7 \\
%v6866603 & True      & 1                         & 23  & 2.8 \\
%         & False     & 1                         & 36  & 3.0 \\
\end{tabular}
\caption{Total number of {\tt DiaSources} detected at 5--sigma from
  all 9 sensors in raft 2,2, for the configurations that yielded the
  lowest number of false detections.  Matches are determined using a
  3'' match radius with the input reference catalog.  Objects with the
  {\tt EDGE} flag are ignored.  The final column lists ratios of the
  number of negatively detected false detections to those with
  positive flux. \label{tab-bestfp10b}}
\end{table}

%\begin{table}
%\centering
%\begin{tabular}{clccc}
%\hline
%\multicolumn{5}{|c|}{Best Configs: Production 8} \\
%\hline
%Visit    & Prefilter & nGauss & degGauss & spatialOrder \\
%\hline
%v6866601 & True      & 3      & 6        & 5 \\
%         & False     & 3      & 2        & 4 \\
%v6866602 & True      & 3      & 6        & 5 \\
%         & False     & 3      & 5        & 5 \\
%v6866603 & True      & 3      & 6        & 4 \\
%         & False     & 3      & 6        & 4 \\
%\end{tabular}
%\caption{Configurations that led to the best results listed in
%  Table~\ref{tab-bestfp8}.  All post-production runs include
%  perturbations about these configurations. \label{tab-bestconfig8}}
%\end{table}
